---
title: "mixOmics with caret examples"
author: "Jonathan Thorsen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This is an example of how we can use sparse Partial Least Squares for modelling. The model is implemented in the mixOmics package (http://mixomics.org/).

First, we load some packages.
```{r, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(mlbench)
data(Sonar)
library(doMC)
library(caret)
library(mixOmics)
library(tidyverse)
library(pROC)
library(mixOmicsCaret)  # This package
```

Then we can register doMC for the built-in multithreading in `caret`. A common option is number of threads minus one. On a quad-core with hyperthreading (Macbook Pro i7), that's 7.
```{r}
registerDoMC(cores = 7)
```

We start by examining the dummy data we'll be using. It's called Sonar and comes from the `mlbench` package.
```{r}
head(Sonar)
Sonar$Class
dim(Sonar)
```

Alright, 60 predictors, 208 observations, 2 classes.
Let's partition the data into a training set and a test set.
```{r}
set.seed(123)
inTraining <- createDataPartition(Sonar$Class, p = .75, list = FALSE)
training <- Sonar[ inTraining,]
testing  <- Sonar[-inTraining,]
```

Now we will setup a cross validation to see how many predictors we should include. First, let's set up the options. It will run 5 repeats of 10-fold CV. We will keep the resamples and the predictions, and allow it to run in parallel. We will turn off verbose-mode for now. 
We save this as a trainControl object, which is just a set of options for `caret`.
```{r}
repCV10 <- trainControl(method = "repeatedcv", 
                        number = 10, 
                        repeats = 5, 
                        returnResamp = "all", 
                        savePredictions = "all", 
                        allowParallel = T, 
                        verboseIter = F)
```

Now for the CV loop itself. This will take some time. We have asked it to model Class M as a numeric 0/1, which is fine for sPLS but `caret` throws a warning we can safely ignore. We will try 1 to 5 components, with a range of keepX values. This is the soNote the `fixX` argument, which is left empty. 
```{r}
set.seed(321)
SonarPLS <- train(as.numeric(Class == "M") ~ ., data = training,
                  method = get_mixOmics_sPLS(),
                  preProc = c("center", "scale"),
                  tuneGrid = expand.grid(ncomp = 1:5, keepX = c(1, 2, 4, 8, 16, 25, 40, 50, 60), keepY = 1),
                  trControl = repCV10,
                  fixX = c())
```

Let's plot the results!
```{r, fig.show='hold'}
SonarPLS
ggplot(SonarPLS, metric = "RMSE")
ggplot(SonarPLS, metric = "Rsquared")
```

We can also inspect the various folds for 
SonarPLS %>% getBestPredictions %>% group_by(Rep) %>% summarize(auc = pROC::auc(obs, pred))%>% arrange(auc)

SonarPLS$pred %>%
  separate(Resample, c("Fold", "Rep")) %>%
  group_by(ncomp, keepX, Rep) %>%
  summarize(auc = pROC::auc(predictor = pred, obs)) %>%
  ggplot(aes(x = factor(keepX), y = auc)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(color = factor(ncomp)), alpha = 0.2, show.legend = FALSE) +
  geom_hline(yintercept = 0.5) +
  facet_grid(. ~ ncomp) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  theme_bw() + theme(strip.background = element_blank()) +
  xlab("Number of variables")

auc(predictor = predict(SonarPLS, testing), testing$Class)








## Example stuff

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
